---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```


### Task 3: Use janitor to clean up the column names



### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crimes |> mutate(total_rate = grand_total/population * 10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crimes_with_rates <- crimes |> mutate(violent_rate = violent_crime_total/population * 10000, property_rate = property_crime_total/population * 10000)
```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

All of the years with the lowest violent crime rates were 2020 and in the 2010s. The same is true for property crime. As for the highest crime rates, violent crime appeared to be highest in thw 1990s, with 1992 having an average of 99 violent crimes per 10,000 people. This was the highest violent crime rate. For property crime, the data is a bit less consistent, with five of the 10 years with the highest rates being in the 1970s and 1980s. The other five years were all in the 90s, so the property crime rate data still contributes to the claim that crime per capita has generally decreased in Maryland. This data is generally in line with my expectations, although it was a little surprising to see every year in the 2010s among the lowest crime rates. I didn't think the data would be that drastic.   


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

