---
title: "SoccerFouls"
---

Do fouls matter in women's college soccer? In this assignment, you'll be working with a dataset that contains information about soccer matches from last season. You'll load the data, perform some basic statistical analysis, and create a visualization to explore the relationship between fouls and the goal differential (and thus the outcome). Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the soccer dataset from https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv")
```


### Task 3: Create a new column called `goal_diff` that calculates the difference between the two teams' goals.

```{r}
matches <- matches |> mutate(goal_diff = team_score - opponent_score)
```



### Task 4: Use `summarize()` and the `cor()` function to calculate the correlation between fouls and goal differential, using the pearson method.

```{r}
matches |> summarize(correlation = cor(fouls, goal_diff, method="pearson"))
```



### Task 5: Create a scatterplot of fouls and goal differential, using `geom_point()`

```{r}

matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=fouls)) +
  geom_smooth(aes(x=goal_diff, y=fouls), method="lm")

```



### Task 6: Describe the results of both the correlation calculation and the scatterplot.
Since the line of best fit is virtually flat, we can conclude that there is little to no correlation between the number of fouls a team commits and the goal differential. Based on the scatterplot, there doesn't seem to be any effect that fouls have on goal differential. 

### Task 7: Pick another variable and see if they're related to goal differential, describe the result.

```{r}
matches |> summarize(correlation = cor(corners, goal_diff, method="pearson"))
```
```{r}
matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=corners)) +
  geom_smooth(aes(x=goal_diff, y=corners), method="lm")
```
The correlation coefficient is approximately 0.378, which would indicate there is a weak positive correlation between the number of corners and goal differential. Corner kicks appears to effect the outcome of the game, but it isn't a dominant factor. 

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
