---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Daniel Stein"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

The Maryland data appears to have a bit more variability than the Pennsylvania data, but more importantly there doesn't appear to be any outliers like with the Pennsylvania data. There are no schools with an 100% FARMS rate and no data points that are really far away from the next closest point.

### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

The Pennsylvania data has a weak positive relationship, compared to the Maryland data, which has a stronger positive relationship. The points of the Maryland data are much less spread out, with more of a stronger pattern than the Pennsylvania data. Even though both counties are suburban, the difference in state FARMS policies could be a significant factor in why the county data looks so different. For example, Pennsylvania might have a more generous policy surrounding free and reduced meals, so there may be more children affected by the policy.


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

There is a strong, negative correlation between poverty and math scores in Montgomery County, MD. According to the model, the higher FARMS percentage, the fewer students meet or exceeded math expectations. The negative coefficient value indicates that there is a negative relationship between the two variables, while the super small p-value tells us that the outcome is statistically significant. Additionally, as a result of a high adjusted R-squared, I think this model is worth using and fits the data better than the Pennsylvania model. 


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

The school with the highest positive residual is Fox Chapel Elementary. Within its zip code, the per capita income is less than the DMV average, as well as the median household income. The areas is also very ethnically diverse, with white people only making up 22% of the population. There isn't anything in the profile that explains why the school is over-performing. If anything, based on the lower-than-average income, I would guess that the school should be under-performing. My first reporting question to the school's principal would be why they think the school is over-performing on math exams. 

The school with the largest negative residual is Germantown Elementary, which also has a median household income and per capita income that are slightly below average in the DMV. The area is slightly less diverse than the zip code for Fox Chapel Elementary, with black and white people making up roughly the same percentage of the population, as well as smaller Asian and Hispanic communities. Something that might explain why the school is under-performing could be that about 37% of the population is foreign-born, which is more than 2x the rate in Maryland (16.6). My first question to the school's principal would be how would they characterize the community of the school. 


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

According to our model, the Maryland data does support the school board member's claim that elementary schools with fewer students on FARMS do better on standardized math tests. This model is much better suited for the Maryland data than the Pennsylvania data, which was much more scattered and revealed a weaker correlation. If I called the school board member, I would ask them what data backs up their claim about FARMS percentage and math test scores. I would also ask him if he supports FARMS programs, in general. In my opinion, the most interesting finding from my analysis is how Fox Chapel Elementary is performing nearly 30 points better than the model predicted. I would visit Fox Chapel first to investigate why its scores are so much higher. I also found interesting the handful of schools in Montgomery County, PA that had a FARMS percentage of 100. 
