---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE
One thing that stands out to me is how there are over 400 people on the White House payroll. I don't know if that is a small or large number, but I think it would be worth comparing to other administrations. I also noticed how there are 8 employees that have a reported salary of 0. That makes slightly more difficult to analyze the data, as those numbers will affect the mean and median salary of the White House officials. When I see that there are people with an offical salary of 0, I wonder what these people do within the White House. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(
  mean(SALARY),
  median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(
  mean(SALARY),
  median(SALARY),
  min(SALARY),
  max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE
Given the fact that the mean is larger than the median, this means the data distribution is not normal, but rather skewed right. The maximum and minimum salaries also tell me the range of the data, which is over $225,000. However, I would argue that the 8 employees with a salary of 0 are outliers because they are $50,000 away from the next highest salary.  
### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE
I think the best story idea is identifying the responsibilities of the the 8 employees with a salary of zero. As the next step, I would try to arrange interviews with them or ask for a statement from the White House on why those 8 employees don't have a salary. 
