---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 * 1000, violent_rate_2020 = violent2020/pop2020 * 1000, property_rate_2020 = property2020/pop2020 * 1000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE
Both in 2019 and 2020, some cities are relatively consistent in how they rank in terms of crime. Baltimore City had the most violent crime in 2019, with around 18 incidents per 1000 people and have the 2nd most in 2020, although its rate decreased to 15 incidents per 1000. Ocean City has by far the most property crime, with 30 more incidents per capita than 2nd-placed Oakland in 2019. In 2020, Ocean City had 64 more property-related reports than 2nd-placed Colmar Manor. Property crime tends to take place in generally more populated cities, whereas violent crime holds less of a correlation. I think the story worth looking into is why Ocean City has so much property crime compared to the rest of Maryland. In both 2019 and 2020, it easily has the most incidents per capita and the amount of crimes increased between 2019 and 2020. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(violent_pct_change = (violent_rate_2020 - violent_rate_2019)/violent_rate_2019 * 100, property_pct_change = (property_rate_2020 - property_rate_2019)/property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE
I think the story in this data is violent crime in ocean city increasing by 82% and property crime increasing by 16% between 2019 and 2020. This story is especially interesting considering 2020 was the beginning of the pandemic, which might cause assumptions that crime would decrease because of people staying inside. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Violent crime in Ocean City increased by 82% between 2019 and 2020. 
According to a study of Maryland crime data, Ocean City ranked in the top 20 among highest increases in violent crime. In 2019, there were 91 reported incidents of violent crime, compared to 166 in 2020. What makes this stat particularly alarming is the increase happening in a larger town like Ocean City. Compared to the 19 other cities with steeper violent crime increases, Ocean City is one of only two cities with a 2020 population greater than 6000. 10 of the cities in the top 20 have populations of 1500 or less, speaking to the unusually large increase in violent crime in a city as populated as Ocean City.

ANSWER HERE

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Violent Crime Percentiles & IQR

````{r}
md_cities_percentiles <- md_cities_change |> mutate(percentile = ntile(violent2020, 100))
````
There are 10 cities above the 90th percentile. 

````{r}
md_cities_percentiles |> summarize(violent2020_IQR = IQR(violent_rate_2020))
````
I don't think finding out the IQR changes my best story idea, as the IQR is not a terribly large number. Because the IQR is fairly small, it means that the middle 50% of the data doesn't have that much variability. The IQR doesn't affect my original idea of looking into Ocean City, as the town had one of the highest violent crime rates. 
