---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
MandMs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```
### Task 3: Sample 5 rows of data from the dataset
```{r}
set.seed(123)

sample_5 <- MandMs |> sample_n(5)
```
### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms
```{r}
sample_5 |> summarize(mean = mean(red), sd = sd(red))
```
### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4
```{r}
MandMs |> summarize(mean = mean(red), sd = sd(red))
```
### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?
The sample mean is 22 M&Ms larger than the population mean. There is not as big of a difference between the two standard deviations, as the sample SD is 4 less than the population standard deviation.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

This sample tells us that the M&Ms dataset could be skewed, as the sample gave me a much larger mean than the population. I think that if I repeated this sample or if I sampled a larger amount of rows, my sample mean and standard deviation would be closer to the population mean and SD. This experiment also tells us that random sampling can sometimes produce unusual results. In the event that a random sample is far off from the population mean and standard deviation, it may be helpful to re-run or change the sample to ensure your data is accurate. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
