---
title: "HW9: Car Theft Confidence Intervals & Margin of Error"
name: Daniel Stein
---

Your editor wants to know what a "typical" month of car thefts looks like in Prince George's County. Not just the average — she wants to know how confident you can be in that number. Using monthly figures, confidence intervals and margin of error, you'll come up with an answer and decide whether the data supports a clear story. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# LOAD THE TIDYVERSE
library(tidyverse)
```

### Task 2: Load the Prince George's county crime data

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. What's your initial reaction to the change in car thefts? Is there a trend? If your editor asked you "how many car thefts happen in a typical month?", what would you say based on this chart alone?

My initial reaction to the change in car thefts is that car thefts have been decreasing over the last few years. While there are a couple spikes in the number of car thefts, the trend seems to be a decrease in car thefts. If my editor asked me how many car thefts happen in a typical month, I would probably say around 400. 

```{r}

car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2026)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```

### Task 4: Calculate the mean, standard error and margin of error

We've got a small number of months, so we need to quantify the uncertainty around our average. We'll calculate the standard error (which measures how much the sample mean could vary), and the margin of error (which gives us the plus/minus range). Replace the REPLACE_ME values based on what we did in class.

```{r}

theft_summary <- car_thefts |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts
  )

# Display the results
theft_summary

```

**Answer this question:** Explain in your own words what the margin of error tells us here. If you were writing a story, how would you use this number? For example, how would you complete this sentence: "Prince George's County averages about ___ car thefts per month, plus or minus ___."

The margin of error tells us that the mean could be 34.18687 larger or smaller than what it currently is (375.5625). If I were writing a story, I would explain that Prince George's County averages 375 car thefts per month, plus or minus 34. I think that I would still use the mean of 375 as an average, but I would caveat it with the margin of error. 

### Task 5: Build the confidence interval

Now let's turn the margin of error into a full confidence interval — the range where we're 95% confident the true average falls. Replace the REPLACE_ME values.

```{r}

theft_summary <- theft_summary |>
  mutate(
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

theft_summary |>
  select(mean_thefts, margin_error, ci_lower, ci_upper)

```

**Answer this question:** If a county official claimed "car thefts are down to about 300 per month," does your confidence interval support that claim? What about a claim of 350 per month? Explain how the confidence interval helps you evaluate each claim.

According to our confidence interval, the county official would be incorrect because we're 95% confident that the true mean is between 341 and 409. If he claimed car thefts were closer to 350 per month, our confidence interval contains 350, so he would be supported by the data.

### Task 6: Visualize the data with the confidence interval

Let's make a chart showing the data with the mean and the 95% confidence interval. Does this change anything about your answer from Task 3?

Adding the calculated to the chart does change my answer from Task 3, as intiailly I estimated that the mean is roughly 400. Now according to the adjusted chart, the mean is around 375.

```{r}
# Make a chart with the mean and CI
car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper),
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month),
           y = theft_summary$mean_thefts + 10,
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)),
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 7: The margin of error and sample size

In class we learned that MoE = 1/√n × 100 as a quick approximation for polls. That same principle applies here: more data means a smaller margin of error. Let's see what would happen if we had 40 months of data instead of the number we have, assuming the same mean and standard deviation. Replace the REPLACE_ME value with 40.

```{r}

hypothetical_summary <- tibble(
  mean_thefts = theft_summary$mean_thefts,
  sd_thefts = theft_summary$sd_thefts,
  months = 40,
  se_thefts = sd_thefts / sqrt(months),
  margin_error = qt(0.975, df = months - 1) * se_thefts,
  ci_lower = mean_thefts - margin_error,
  ci_upper = mean_thefts + margin_error
)

# Compare the two
comparison <- tibble(
  scenario = c("Actual data", "40 months"),
  months = c(theft_summary$months, hypothetical_summary$months),
  margin_error = c(theft_summary$margin_error, hypothetical_summary$margin_error),
  ci_lower = c(theft_summary$ci_lower, hypothetical_summary$ci_lower),
  ci_upper = c(theft_summary$ci_upper, hypothetical_summary$ci_upper)
)

comparison

```

**Answer this question:** How did the margin of error and confidence interval change? What does this tell you about the tradeoff between having more data and being able to make precise claims? If you were covering crime trends, how many months of data would you want before writing a definitive story?

As the amount of data increased, the margin of error decreased and the range of the confidence interval increased. Having more data means that you are not able to make as precise claims, especially without changing your level of confidence. While it depends on the story angle, if I was covering crime trends, I would ideally want multiple years of crime data, so at least 24 months. However, it depends on what my story exactly is. 

### Task 8: Reporting with uncertainty

Based on everything you've calculated, answer the following:

1. Write a headline and a one-paragraph lede about car thefts in Prince George's County. Your text should properly reflect the uncertainty in the data — don't overstate or understate what the numbers show.

2. Your editor says "Just tell me if car thefts are going up or down." Using only the confidence interval (no significance testing yet), what's the most accurate answer you can give? What would you need to do next to answer that question more definitively?

Headline: Car thefts in Prince George's County have decreased since 2023, county still averages over 300 per month.

According to data from Prince George's County, the number of monthly car thefts has decreased by 400 since 2023. However, the mean number of monthly car thefts remains over 300, and the county could average as many as 400 thefts per month. The dataset begins in July 2023, when PG County had over 600 car thefts in one month. By the end of the dataset, February 2026, the county had just over 200.

The most accurate answer I could give my editor is that there is a 95% chance that the mean number of car thefts is between 341 and 409 with a margin of error of 30. Another step to answer the question more definitively is to collect more data and try to identify a trend from a longer period of time.  

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
